---
title: "STAT645 - Homework 2"
author: "Salih Kilicli"
date: "9/3/2019"
output:
  html_document: default
  word_document: default
  pdf_document: 
    latex_engine: lualatex
editor_options:
  chunk_output_type: inline
---

**Problem 1: With the calcium data in “calcium.txt,” consider the Decrease variable as your response
and Treatment as your treatment. In what follows, I have recoded Treatment to equal 0 for
placebo and 1 for calcium treatment.**

**(a) For the regression model $Decrease_i = \beta_0 + \beta_1 Treatment_i + \epsilon_i$, write down the model matrix.**

The model can be defined as $Y=X \beta + \epsilon$ where the model matrix, $X$, is given by:

\[X=\begin{bmatrix}
1&1\\
\vdots&\vdots\\
1&1\\
1&0\\
1&0\\
\vdots&\vdots\\
1&0\\
\end{bmatrix}_{21x2}, \qquad Y=\begin{bmatrix}
y_1\\
\vdots\\
\vdots\\
\vdots\\
\vdots\\
y_{21}\\
\end{bmatrix}_{21x1}, \qquad \epsilon=\begin{bmatrix}
e_1\\
\vdots\\
\vdots\\
\vdots\\
\vdots\\
e_{21}\\
\end{bmatrix}_{21x1}, \qquad \beta=\begin{bmatrix}
\beta_1\\
\beta_2\
\end{bmatrix}_{2x1}\]

Notice, above $X$ is a $21x2$ matrix whose second column consists of 10 ones and 11 zeros representing calcium treatment and placebo treatment, respectively.


**(b) Fit the above model, and report the coefficient estimates and standard errors.**

```{r}
setwd("~/Desktop/STAT645/Data")
calc <- read.table("calcium.txt", header=TRUE)
model1 = lm(Decrease ~ Treatment, data = calc)
summary(model1)
```

**(c) Based on the model, what is the p-value for the null hypothesis of no treatment effect?**

```{r}
p = summary(model1)[["coefficients"]][2,4]
cat("p-value for the null hypothesis of no treatment effect is p =", p)
```

**(d) Now analyze the same data using a two-sample t-test, assuming equal variances. How
do the results compare to those you obtained using the regression model?**

```{r}

```

**(e) Assuming that the $\epsilon_i$ are normally distributed, what is the estimated distribution of
$Decrease$ when $Treatment = 1$?**

```{r}

```


**Problem 2: With the onset data in “onset data.csv,” conduct the following analysis.**

**(a) Create side-by-side box plots comparing time to onset with (i) the tx variable and (ii)
the prior variable. Comment.**

```{r}
setwd("~/Desktop/STAT645/Data")
ons = read.csv("onset_data.csv", header=TRUE)
par(mfrow = c(1,2))
boxplot(onset ~ tx, data = ons, col = rainbow(2), las=1)
boxplot(onset ~ prior, data = ons, col = rainbow(2), las=1)
```

**(b) Create a scatterplot of onset vs. age. Color code the points by prior status. Also, fit
and overlay separate lowess curves, one each for prior = 0 and prior = 1.**

```{r}
attach(ons)
plot(onset, age, col = rainbow(2), las=1)
lines(lowess(onset,age), col="purple")
lines(lowess(onset,age+prior), col="green")
```

**(c) Fit the regression model
$$y_i = \beta_0 + \beta_1 tx_i + \beta_2 prior_i + \beta_3 age_i + \beta_4 (prior × age)_i + \epsilon_i$$
Interpret all coefficients and report their estimates and standard errors.**

```{r}
model2 = lm(onset ~ tx + prior + age + I(prior*age))
summary(model2)
```

**(d) Use matrix manipulation using a design matrix to verify the estimates and standard
errors from above.**

**(e) What is a 95% confidence interval for the mean difference in onset times between the
treatment and control groups, holding prior status and age constant?**

**(f) What is a 95% confidence interval for the mean response of a treated individual, age 35,
with no prior tumor incidence?**

**3. Suppose that $y_1, y_2, . . . , y_n$ are i.i.d. realizations from the $N(0, \sigma^2)$ distribution. Derive the
maximum likelihood estimator of $\sigma^2$.**

```{r}
 
```

**Problem 4: Suppose the times to infection following exposure to a particular bacteria follow the gamma
distribution with shape parameter $\alpha$, scale parameter $\beta$, and pdf 
$$f(x) = \dfrac{1}{\Gamma(\alpha)\beta^{\alpha}}x^{\alpha-1}e^{-x/\beta}$$
Use the *nlm* function in R to compute the maximum likelihood estimates for the data in
“gamma.csv.” **

```{r}

```
