---
title: "STAT645 - Homework 9"
author: "Salih Kilicli"
date: "10/1/2019"
output:
  html_document:
    df_print: kable
    highlight: zenburn
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---

```{r}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
```

**Problem 1: Consider the monoclonal gammopathy of undetermined significance (MGUS) dataset available under survival package. Consider age, gender, albumin, creatinine, and hemoglobin level at MGUS diagnosis as the potential covariates. Work with the dataset where these covariates are completely observed. Fit a proportional hazard (PH) model for the days-to-death from the diagnosis in terms of the covariates.**

```{r}
library(survival)
mydata0=data.frame(mgus$age, mgus$sex, mgus$alb, mgus$creat, mgus$hgb, mgus$death, mgus$futime)
mydata=mydata0[complete.cases(mydata0), ]
out=coxph(Surv(mgus.futime, mgus.death)~mgus.age+mgus.sex+mgus.alb+mgus.creat+mgus.hgb, data=mydata)
```

**(a) Test the association between the hazard and the age of diagnosis at the $5\%$ level.**

Null and Alternative hypotheses for testing the association between the hazard and the age of diagnosis can be written as:

$H_0:$ There is no association between proportonal hazard and the age of diagnosis

$H_a:$ There is association between proportonal hazard and the age of diagnosis 

```{r}
p1=summary(out)
p1
```

Test statistic for the testing $H_0$ at the $5\%$ is given by $Z=8.223$ and the p-value is $p_{age} <2\times e^{-16}<0.05$. Therefore, we reject the Null hypothesis at $5\%$ level and there is a statistically significant evidence that there is association between hazard and the age of diagnosis.


**(b) Estimate relative risk and its $95\%$ CI for the death of a subject with the age of diagnosis $60$ compared to the subject with the age of diagnosis $50$ while all other covariates remain unchanged.**

Let $Z$ denotes the covariates other than $mgus.age$. Then, relative risk can be written as:

$$RRisk = \dfrac{\lambda(t|mgus.age=60,Z=Z_0)}{\lambda(t|mgus.age=50,Z=Z_0)}=\dfrac{\lambda_0(t)exp(60\beta_{mgus.age}+Z_0^T\gamma)}{\lambda_0(t)exp(60\beta_{mgus.age}+Z_0^T\gamma)}=exp(10\beta_{mgus.age})=\big(exp(\beta_{mgus.age})\big)^{10}$$
And, the summary already gives estimates and $\%95$ CI for unit change in the variables. Therefore, simply raising the answer to the $10$th power yields the estimate and $\%95$ CI for the problem. Same values can be found by using estimates $\hat{\beta}_{mgus.age}$ and $se(\hat{\beta}_{mgus.age})$ such as:

estimate = $exp(10*\hat{\beta}_{mgus.age})$

CI = $exp(10*[\hat{\beta}_{mgus.age}\pm 1.96*se(\hat{\beta}_{mgus.age})])$

```{r}
((p1$conf.int[1,])^10)[-2]   #this gives you the estimates to the power 10th exp(10B)=(exp(B))^10

# or we can calculate it by hand

Estimate = (p1$conf.int[1,1])^10
Lower = (p1$conf.int[1,3])^10
Upper = (p1$conf.int[1,4])^10
CI = cbind(Lower, Estimate, Upper)
CI
```

**(c) Test if there is any effect of gender, albumin, and hemoglobin at the $5\%$ level [This is a composite hypothesis test, so do not test them separately].**

```{r}
out0=coxph(Surv(mgus.futime, mgus.death)~mgus.age+mgus.creat, data=mydata)
outa=coxph(Surv(mgus.futime, mgus.death)~mgus.age+mgus.sex+mgus.alb+mgus.creat+mgus.hgb, data=mydata)
anova(out0,outa)
```

$$H_0: (\beta_{mgus.sex},\beta_{mgus.alb},\beta_{mgus.hgb})=(0,0,0)$$

$$H_a:  (\beta_{mgus.sex},\beta_{mgus.alb},\beta_{mgus.hgb}) \neq (0,0,0)$$
The Chi-sqrd test statistics is $7.3248$ and the p-value is $p = 0.06224 > 0.05$. Therefore, we fail to reject the Null hypothesis and conclude that there is $\textbf{no}$ significant effect of the covariates $mgus.sex$,$mgus.alb$,$mgus.hgb$ on days-to-death from the diagnosis.

**(d) Based on the initial PH model with age, gender, albumin, creatinine, and hemoglobin as the covariates, obtain the estimate and $95\%$ CI for the $10$ year survival probability for subjects with the following covariate values.**

Age | Gender | Albumin | Creatinine | Hemoglobin
----| ------ | ------- | ---------- | ----------
60  | Male   |    3    |      1     |    13.5    
60  | Male   |    3    |      4     |    13.5  

```{r}
creat = c(1,4)
estimate = lower = upper = matrix(c(0,0), nrow=2)

for (i in 1:2){
out1d=survfit(out, newdata=data.frame(mgus.age = 60, mgus.sex = "male", mgus.alb = 3, 
                                      mgus.creat = creat[i], mgus.hgb = 13.5))
index1=findInterval(10*365, out1d$time)
estimate[i,] = out1d$surv[index1]      # estimate of S(10 years|given the covariate values)
lower[i,] = out1d$lower[index1]
upper[i,] = out1d$upper[index1]        # the 95% CI
}

CI=cbind(lower, estimate, upper)
rownames(CI) = c("First","Second")
colnames(CI) = c("Lower", "Estimate", "Upper")
CI
```

**(e)  Suppose now all two factor interactions among the covariates age, gender, albumin, creatinine, and hemoglobin are now included in the model along with the main effects. Apply a stepwise model selection technique to choose the best fitted model. Now based on the best fitted model obtain the estimate and $95\%$ CI for the $10$ year survival probability for subjects with covariate values mentioned in the previous question.**

```{r}
mydata0=data.frame(mgus$age, mgus$sex, mgus$alb, mgus$creat, mgus$hgb, mgus$death, mgus$futime)
mydata=mydata0[complete.cases(mydata0), ]
out5=coxph(Surv(mgus.futime, mgus.death)~ mgus.age + mgus.age*mgus.sex+mgus.age*mgus.alb+mgus.age*mgus.creat+mgus.age*mgus.hgb
                                        + mgus.sex + mgus.sex*mgus.alb + mgus.sex*mgus.creat + mgus.sex*mgus.hgb
                                        + mgus.alb + mgus.alb*mgus.creat + mgus.alb*mgus.hgb
                                        + mgus.creat + mgus.creat*mgus.hgb
                                        + mgus.hgb, data=mydata)
oute = step(out5, direction="both", trace=0)
summary(oute)

out6=coxph(Surv(mgus.futime, mgus.death)~ mgus.age 
                                        + mgus.sex + mgus.sex*mgus.alb 
                                        + mgus.alb + mgus.alb*mgus.hgb
                                        + mgus.creat + mgus.creat*mgus.hgb
                                        + mgus.hgb, data=mydata)

creat = c(1,4)
estimate = lower = upper = matrix(c(0,0), nrow=2)
for (i in 1:2){
out1d=survfit(out6, newdata=data.frame(mgus.age = 60, mgus.sex = "male", mgus.alb = 3, 
                                      mgus.creat = creat[i], mgus.hgb = 13.5))
index1=findInterval(10*365, out1d$time)
estimate[i,] = out1d$surv[index1]      # estimate of S(10 years|given the covariate values)
lower[i,] = out1d$lower[index1]
upper[i,] = out1d$upper[index1]        # the 95% CI
}
CI=cbind(lower, estimate, upper)
rownames(CI) = c("First","Second")
colnames(CI) = c("Lower", "Estimate", "Upper")
CI
```

**Problem 2: Suppose that a two-arm clinical trial is being designed to test if a new drug is equivalent to the existing drug in delaying terminally ill liver cancer patients. Suppose that two drugs will be considered to be equivalent if the difference in the proportions of subjects who survive at least $3$ years after the surgery does not exceed $8\%$ (0.08) (FDA requires $\pm 20\%$ of the overall average as the equivalence limit). Each group received 40 subjects, the sample proportions for the existing and new treatment are $40\%$ and $45\%$. Test the null hypothesis of non-equivalence at the $5\%$ level.**

The Null and Alternative hypotheses for the testing of equivalence of the drugs can be written as:

$$H_0: |\pi_1-\pi_2| \geq 0.08 \quad (i.e., \quad (\pi_1-\pi_2)\leq -0.08 \quad \text{or} \quad (\pi_1-\pi_2) \geq 0.08)$$
$$H_a: |\pi_1-\pi_2| < 0.08 \quad (i.e., \quad -0.08 < \pi_1-\pi_2 < 0.08)$$

```{r}
m=n=40; pihat1 = 0.45; pihat2 = 0.40
se = sqrt((pihat1*(1-pihat1)+pihat2*(1-pihat2))/40)
t = qnorm(0.025,lower.tail = FALSE)
CI = (pihat1-pihat2) + c(-1,1)*t*se
names(CI)=c("Lower", "Upper")
CI
```

Since the $95\%$ level confidence interval has values outside of the $-0.08 < \pi_1-\pi_2 < 0.08$ interval (ex: 0.25 > 0.08), we fail to reject the Null hypothesis. Thus, we conclude that there is no enough evidence that the two drugs are equivalent.

**Problem 3: Suppose that in a two-arm clinical superiority trial we have obtained the following dataset. Analyze the data and make your conclusions.**

<br/>     | Control | Treatment |
--------- | ------  | --------- |
Success   |    10   |     14    |  
Failure   |    12   |     9     |
Total     |    22   |     23    |
Withdrawn |     3   |     5     |

Analysis of the data can be done in two different methods: Per protocol analysis vs Intention-to-treat analysis. Using Intent-to-Treat Analysis (considering withdrawns as Failure) we get the table below for failure and success probabilities:

Intent-to-Treat Analysis | Control | Treatment |
------------------------ | ------  | --------- |
P(Success)               |  $40\%$ |   $50\%$  |  
P(Failure)               |  $60\%$ |   $50\%$  |
Total                    |    22   |     23    |
Withdrawn                |     3   |     5     |

Null and Alternative hypotheses for the superiority trial can be written as:

$$H_0: (\pi_1-\pi_2) = 0$$
$$H_a: (\pi_1-\pi_2) > 0$$

```{r}
m=28; n=25;
pi_t = 0.50; pi_c = 0.40
se = sqrt(pi_t*(1-pi_t)/m+pi_c*(1-pi_c)/n)
t = qnorm(0.05, lower.tail=FALSE)           #one-sided alternative used to find t-statistics
CI = (pi_t-pi_c) + c(-1,1)*t*se
names(CI)=c("Lower", "Upper")
CI
```

Since the confidence interval found contains $0$ (or have negative values) we fail to reject the Null hypothesis at $5\%$ level and conclude that there is no statistically enough evidence that Treatment group is superior to Control  group.


**Problem 4: Following data were collected from a cross-over trial of comparative effectiveness of medical interventions for ocular hypertension and open-angle glaucoma. Analyze the data, check for the treatment effect and justify your method(s).**

Group 1 (A then B) | Period (1    2) |
-------------------|-----------------|
1                  | 30  _______  20 |
2                  | 32  _______  19 |
3                  | 28  _______  20 |
4                  | 32  _______  24 |
5                  | 31  _______  22 |

Group 2 (A then B) | Period (1   2 ) |
-------------------|-----------------|
6                  | 22  _______  30 |
7                  | 23  _______  29 |
8                  | 20  _______  31 |
9                  | 25  _______  32 |
10                 | 21  _______  28 |

```{r}
library(lme4)

id = rep(1:10, each=2)
period = rep(c(1:2), 10)
treatment=c(rep(c("A","B"),5),rep(c("B","A"),5))
response = c(c(30,20,32,19,28,20,32,24,31,22),c(22,30,23,29,20,31,25,32,21,28))
period = as.factor(period)
xoverdata=data.frame(id, period, treatment, response)
head(xoverdata)
tail(xoverdata)

outp4 = lmer(response~period+treatment+treatment*(period)+(1|id))
summary(outp4)
outp4_0 = lmer(response~period+treatment+(1|id))
anova(outp4_0, outp4)
```

We obtain $\hat{\sigma}_{\tau}^2=1.25$ that is less than $\hat{\sigma}_e^2=2$. The interaction effect  turns out to be statistically non-significant with a large p-value $p=0.72$. Also, period seems to have no effect on the mean of the response variable. Therefore, we shall fit the model without any interaction.

```{r}
summary(outp4_0)
```

The results indicate that there is a statistically significant effect of the treatment. Compared to group A, the mean comparative effectiveness of medical interventions is 8.700 unit lesser in the group B. Also, since the period effect -0.900 that means average value of the response variable will be change by 0.900 units if you take observation from Period 1 to Period 2.

```{r}
anova(outp4_0)
```